---
title: "Covid data analysis - Africa"
author: "Isha Doshi"
date: "2022-10-21"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Loaded a single month of African data. Loaded the list of African countries countries-africa.csv. There are 58 countries listed here.**
```{r}
Africa <- read.delim("covid/countries-africa.csv.bz2")
head(Africa,5)
nrow(Africa)
any(is.na(Africa$country)) 
# checking if any country is NA. There are 58 African countries listed in the database.
```

**Collected all the names of covid data files into a character vector. There are 21 files with names starting with covid data.**

```{r}
vec= list.files(path = "covid/", pattern = "^covid-global", all.files = FALSE,
           full.names = FALSE, recursive = FALSE,
           ignore.case = FALSE, include.dirs = FALSE, no.. = FALSE)
vec
#list.files produces a character vector so we can find the length directly.
length(vec)
```

**Loaded the COVID data file for October 2021.Since the global data file contains not just African countries, I just selected the African ones from the list. Unfortunately not all the names match.**

```{r}
library(tidyverse)
library (dplyr)
october2021= read.delim("covid/covid-global_10-01-2021.csv.bz2")
head(october2021)
# checking the names of the columns in Oct 2021 data
names(october2021)
```

**49 countries can be found in Oct 2021 covid data from African countries list**
```{r}
library(tidyverse)
library (dplyr)
# filtering oct data based on African countries 

AfricaOct2021<- october2021[october2021$Country_Region %in% Africa$country,]

# tried two methods 

AfricaOct2021 <- filter(october2021,october2021$Country_Region %in% Africa$country)
head(AfricaOct2021)

nrow(AfricaOct2021)

```
**The 9 countries below are not matched in covid data.**
```{r}
`%notin%` <- Negate(`%in%`)
MissingAfricanCountries <- filter(Africa,Africa$country %notin% october2021$Country_Region)
MissingAfricanCountries
```

**We should care more about these countries as their population is higher than the smaller islands. If these countries are ignored, the number of cases and deaths due to covid in Africa could be miscalculated entirely. The data we would be using further would not stay accurate.**

**Next, find how are the names of these three countries (Two Congos and Ivory Coast) written in the covid data. After looking through the Oct 2021 data's list of countries, I found that these 3 countries are written as Congo (Kinshasa), Congo (Brazzaville) and Cote d’Ivoire.**

```{r}
names<-october2021 %>% filter(str_detect(october2021$Country_Region, "^Congo|^Cote"))
names
```

**I replaced Ivory Coast in the list of African countries with Cote d'Ivoire, Democratic Republic of the Congo with Congo (Kinshasa) and Republic of the Congo with Congo (Brazzaville).**
```{r}
Africa$country[Africa$country =='Ivory Coast']<-"Cote d'Ivoire"
Africa$country[Africa$country=='Democratic Republic of the Congo']<-"Congo (Kinshasa)"
Africa$country[Africa$country=='Republic of the Congo']<-"Congo (Brazzaville)"
head(Africa)

# I am only left with “Réunion (France)”, “Western Sahara”, “Cape Verde”, “Mayotte (France)”, “São Tomé and Príncipe”, and “Saint Helena, Ascension and Tristan da Cunha (UK)”.

`%notin%` <- Negate(`%in%`)
MissingAfricanCountries <- filter(Africa,Africa$country %notin% october2021$Country_Region)
MissingAfricanCountries
```

**The file name is written as “covid-global_<mm>-<date>-<yyyy>.csv.bz2”, and date always “01” in these files. Extracting the date part from the first file name as Date object.**

Answer: Removing the date part of file "covid-global_01-01-2021.csv.bz2" and storing it in a object of date type. 
```{r}
filename="covid-global_01-01-2021.csv.bz2"
filename=gsub("^[^_]*_","", filename)
filename=gsub("\\..*","",filename)
filename
firstDate= as.Date(filename, "%m-%d-%Y")
firstDate
class(firstDate)
```

**Now I merge all the data files into one.** 

```{r}
library(tidyverse)
library(lubridate)
#creating a merged df by binding data from each file after performing processes to extract date, select and rename columns, and selecting African countries
mergedDf<-NULL
for (item in vec) {
  df <- read.delim(paste0("covid/", item))
 item=gsub("^[^_]*_","", item)
 item=gsub("\\..*","",item)
 firstDate= as.Date(item, "%m-%d-%Y")
  df$year=year(firstDate)
  df$month=month(firstDate)
  selectedDf=df%>%select(starts_with("Country"),Deaths, year, month)
  selectedDf=selectedDf%>%rename(country=starts_with("Country"))
  mergedDf=rbind(mergedDf,selectedDf)
}
mergedDf<- mergedDf[mergedDf$country %in% Africa$country,]
head(mergedDf)
nrow(mergedDf)
```
**Extracted the population size from the dataset of African countries.**

```{r}
AfricaPopulationDf= Africa%>%select(country,population)
head(AfricaPopulationDf)
```

**For each country, computed the death rate: number of deaths per 1M population.**

```{r}
#Merging AfricanPopulationDf with Africa's covid data
combinedDf=merge(mergedDf,AfricaPopulationDf,by.x="country", by.y="country", all=TRUE)
combinedDf$population <- as.numeric(gsub(",","",combinedDf$population))
combinedDf$deathRate<-(combinedDf$Deaths*100000)/as.numeric(combinedDf$population)
head(combinedDf)
```

**Analyzing which 10 countries have the largest death rate? (As of the latest date in the data, Oct 1st, 2021).**
```{r}
topTen=combinedDf[order(combinedDf$deathRate, decreasing = TRUE), ]
topTen=topTen%>%filter(topTen$year=="2021" & topTen$month=="10")
head(topTen,10)
topTenCountries=head(topTen$country,10)
topTenCountries
```

**Plotting the growth in death rate in these 10 countries over time. **

```{r}
dfToVisualize= combinedDf[combinedDf$country %in% topTenCountries,]
dfToVisualize=within(dfToVisualize, Date <- sprintf("%d-%02d", year, month))
head(dfToVisualize)
ggplot(dfToVisualize, aes(x=Date, y=deathRate, fill=country, group=country)) + geom_point(aes(color = country)) +geom_line(aes(color=country))
```

**Computing the number of new monthly deaths (per 1M population) and displaying it on a similar plot.**

```{r}
#correcting the order of data for using lag 
dfToVisualize=dfToVisualize[order(dfToVisualize$country, dfToVisualize$Date),]
#calculating the difference in the deaths per month, grouped by country
dfToVisualize$monthly_Deaths <- ave(dfToVisualize$Deaths, factor(dfToVisualize$country), FUN=function(x) c(NA,diff(x)))
head(dfToVisualize)
# now that we have monthly deaths, we can calculate monthly death rate and append that column to the df.
dfToVisualize$monthlyDeathRate=dfToVisualize$monthly_Deaths*100000/dfToVisualize$population
head(dfToVisualize)
# now we display this data 
ggplot(dfToVisualize, aes(x=Date, y=monthlyDeathRate, fill=country, group=country)) + geom_point(aes(color = country)) +geom_line(aes(color=country))
```

**Namibia experienced the highest peak in the new monthly deaths. It was in 2020-08 (Aug-2020). I can see three waves of covid in the graph.**
